---
title: "Remote Execution"
output: html_document
---

```{r setup, include=FALSE, echo=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

These are examples of how you can use the `mrsdeploy` library from you local machine
to exploit the compute power a remote VM.

###            Logging into Microsoft R Server                 

Use `remoteLogin` to authenticate with R Server using 
the local admin account. Use session = false if the 
remote R intepreter should not be connected.

```
library(mrsdeploy)
remoteLogin("http://localhost:12800", 
            username = "admin", 
            password = "MLads2017!",
            session = TRUE)
```

###            Example: Switching between local and remote sessions                  

Switching between the local command line and the remote command line 
is done using these functions: pause() and resume(). To switch back to
the local R session, type pause(). If you have switched to the local R 
session, you can go back to the remote R session by typing resume().
To terminate the remote R session, type exit at the REMOTE> prompt. 
Also, to terminate the remote session from the local R session, 
type remoteLogout().

```{r}
x <- rnorm(1000)
ls()
# [1] "x"

pause()  #switches the user to the local R session
ls()
# character(0)
```

Generate a diff report between local and remote machine
```{r}
diffLocalRemote()
```

###            Example: Transferring objects between sessions                    

After you have executed some R code locally or remotely, you may want to move 
those objects between sessions. For example, if you have a local R object that you want
to make available to your remote R session, you can use the function putLocalObject(). 
If you want to sync your local and remote workspaces, the functions putLocalWorkspace()
and getRemoteWorkspace() can be used.

```{r}
# Copy remote workspace to local session
getRemoteWorkspace()

# Create a local file
sink("outfile.txt")
cat("hello world")
sink()

# Copy local file to remote machine
putLocalFile("outfile.txt")
listRemoteFiles()
# [[1]]
# [1] "outfile.txt"

# Copy all objects from local R session to remote R session
a <- "character object"
b <- 1:10
putLocalWorkspace()
resume()
ls()
# [1] "a" "b" "x"
```
### Example: Running batch R job on the server 

There are various uses for the server to take advantage of the VM, in addition to running interactively at the `REMOTE>` prompt. A simple case is to take advantage of the remote server to run large time-consuming jobs. For instance, this interation, to compute a regression's leave-one-out r-squared values---

```{r}
    rsqr <- c()
    system.time(
    for (k in 1:nrow(mtcars)) {
    rsqr[k] <- summary(lm(mpg ~ . , data=mtcars[-k,]))$r.squared
    })
    print(summary(rsqr))
```

This can be done the same remotely:

```{r}
    remoteExecute("rsqr <- c()\
    system.time(\
    for (k in 1:nrow(mtcars)) {\
        rsqr[k] <- summary(lm(mpg ~ . , data=mtcars[-k,]))$r.squared\
    })")
```

We'll need to recall the results separately, since only the last value in the remote expression output is printed:

```{r}
    remoteExecute("summary(rsqr)")
```

For larger chunks of code, you can include them in script files, and execute the file remotely by 
use `mrsdeploy::remoteScript("myscript.R")` which is simply a wrapper around `mrsdeploy::remoteExecute("myscript.R", script=TRUE)`, where `myscript.R` is your file found in your _local_  machine's working directory. Since R has a batch mode from the command line, you
could even run a local batch job to invoke a script that ran a remote batch job, and returned a result to the local machine. 

Note that the the `mrsdeploy` library is not needed in the remote script.  So both 
`library(mrsdeploy)` and `install.packages("mrsdeploy")` will generate an error on the remote session.  _If you've included these statements to enable your local script, be sure to remove them if you execute the script remotely, or the script will fail!_ If you want to use the same script in both places, a simple workaround is to avoid making the library call in the script when it runs in the remore session:

	if ( Sys.info()["user"] != "rserve2" ) {
	  library(mrsdeploy)
	}  
 
The ability of `mrsdeploy` to execute a script remotely is just the tip of the iceberg. It also enables moving files and variables back and forth between local and remote.  The `snapshot()` function makes this easy. 

###           Example: R session snapshots                 

If you need a prepared environment for remote script execution that includes
any of the following: R packages, R objects and data files, consider creating 
a snapshot. A snapshot is an image of a remote R session saved to Microsoft R 
Server, which includes:
   - The session's workspace along with the installed R packages
   - Any files and artifacts in the working directory

```{r}
# Create glm model with `mtcars` dataset
carsModel <- glm(formula = am ~ hp + wt, data = mtcars, family = binomial)

# Produce a prediction function that can use the model
manualTransmission <- function(hp, wt) {
  newdata <- data.frame(hp = hp, wt = wt)
  predict(carsModel, newdata, type = "response")
}

# Test function locally by printing results
print(manualTransmission(120, 2.8)) # 0.6418125

putLocalObject(c("carsModel", "manualTransmission"))
snapshot_id <- createSnapshot("my modeling environment")

# Whenever you need the modeling environment, reload the snapshot on the remote session
rm(list=c("carsModel", "manualTransmission"))

loadSnapshot(snapshot_id)  
# Execute an R script remotely
result <- remoteExecute("manualTransmission(120, 2.8)")

cat(result$consoleOutput)
#         1 
# 0.6418125 
```


###  Finally, logout and terminate the remote R session             #

```{r}
exit 
```

