---
title: "Using SparkR with 2013 NYCTaxi Data: Data wrangling, manipulations, modeling, and evaluation"
date: "`r format(Sys.time(), '%B %d, %Y')`"
author: "Algorithms and Data Science & R Server Teams, Microsoft Data Group"
output: 
  html_document: 
    fig_caption: yes
    fig_height: 4
    fig_width: 4
    highlight: haddock
    keep_md: yes
    number_sections: yes
    theme: journal
    toc: yes
    toc_float: yes
runtime: knit
---

<hr>
#Introduction
This Markdown document shows the use of <a href="https://spark.apache.org/docs/latest/sparkr.html" target="_blank">SparkR</a> for data wrangling, manipulation, and creating machine learning models. The data used for this exercise is the public NYC Taxi Trip and Fare data-set (2013, December, ~4 Gb, ~13 million rows) available from: http://www.andresmh.com/nyctaxitrips. Data for this exercise can be downloaded from the public blob (see below). The data can be uploaded to the blob (or other storage) attached to your HDInsight cluster (HDFS) and used as input into the scripts shown here.

We use Spark SQL for many of the data wrangling tasks. For plotting and visualization, small amounts of data from Spark dataframes are transformed to the local data frames.
<hr>
<br>

#Creating spark connection, loading packages
```{r Load Packages, message=FALSE, warning=FALSE, echo=TRUE}
###########################################
# LOAD LIBRARIES FROM SPECIFIED PATH
###########################################
Sys.setenv(YARN_CONF_DIR="/opt/hadoop/current/etc/hadoop", 
           HADOOP_HOME="/opt/hadoop/current", 
           JAVA_HOME = "/usr/lib/jvm/java-1.8.0-openjdk-1.8.0.111-1.b15.el7_2.x86_64",
           SPARK_HOME = "/dsvm/tools/spark/current",
           PATH=paste0(Sys.getenv("PATH"),":/opt/hadoop/current/bin:/dsvm/tools/spark/current/bin"))

.libPaths(c(file.path(Sys.getenv("SPARK_HOME"), "R", "lib"), .libPaths()))
library(SparkR)
library(rmarkdown)
library(knitr)
library(gridExtra)
library(ggplot2)

###########################################
# CREATE SPARK CONTEXT
###########################################
#sparkEnvir <- list(spark.executor.instance = '4', spark.yarn.executor.memoryOverhead = '4000')
sc <- sparkR.session(
  #sparkEnvir = sparkEnvir,
  sparkPackages = "com.databricks:spark-csv_2.10:1.3.0"
)

###########################################
## SPECIFY BASE HDFS DIRECTORY
###########################################
fullDataDir <- "/user/RevoShare/remoteuser/Data"
```
<br>
<hr>


#Reading in files from HDFS (csv or parquet format)
Data for this exercise can be downloaded from the public blob locations below: 
<br>
1. Trip (Csv): http://cdspsparksamples.blob.core.windows.net/data/NYCTaxi/KDD2016/trip_data_12.csv
<br>
2. Fare (Csv): http://cdspsparksamples.blob.core.windows.net/data/NYCTaxi/KDD2016/trip_fare_12.csv
<br>
The data can be uploaded to the blob (or other storage) attached to your HDInsight cluster (HDFS) and used as input into the scripts shown here. The csv files can be read into Spark context and saved in parquet format. Once saved in parquet format, data can be read in much more quickly than csv files.

##Read in data from csv files
```{r Read in files, message=FALSE, warning=FALSE, echo=TRUE}
###########################################
# TRIP FILE (CSV format)
###########################################
tripPath <- file.path(fullDataDir, "trip_data_12.csv")
tripDF <- read.df(tripPath, source = "com.databricks.spark.csv", 
                  header = "true", inferSchema = "true")
#SparkR::cache(tripDF); SparkR::count(tripDF)
head(tripDF, 3)
printSchema(tripDF)

###########################################
# FARE FILE (parquet format)
###########################################
farePath <- file.path(fullDataDir, "trip_fare_12.csv")
fareDF <- read.df(farePath, source = "com.databricks.spark.csv", 
                  header = "true", inferSchema = "true")
#SparkR::cache(fareDF); SparkR::count(fareDF)
head(fareDF, 3)
printSchema(fareDF)
```

#Data wrangling & cleanup using SQL
SparkR is an R package that provides a light-weight frontend to use Apache Spark from R. In Spark 2.0, SparkR provides a distributed data frame implementation that supports operations like selection, filtering, aggregation etc. (similar to R data frames, dplyr) but on large datasets. SparkR also provides support for distributed machine learning using MLlib.

You can register dataframes as tables in SQLContext and join using multiple columns. The following SQL also filters the data for some outliers.
```{r Register tables, message=FALSE, warning=FALSE, echo=TRUE}
###########################################
# 1. REGISTER TABLES AND JOIN ON MULTIPLE COLUMNS, FILTER DATA
# 2. REGISTER JIONED TABLE
###########################################
createOrReplaceTempView(tripDF, "trip")
createOrReplaceTempView(fareDF, "fare")

trip_fareDF <-  SparkR::sql("SELECT 
  f.pickup_datetime, hour(f.pickup_datetime) as pickup_hour, 
  t.dropoff_datetime, hour(t.dropoff_datetime) as dropoff_hour,
  f.vendor_id, f.fare_amount, f.surcharge, f.tolls_amount, 
  f.tip_amount, f.payment_type, t.rate_code, 
  t.passenger_count, t.trip_distance, t.trip_time_in_secs, 
  t.pickup_longitude, t.pickup_latitude, t.dropoff_longitude, 
  t.dropoff_latitude
  FROM trip t, fare f  
  WHERE t.medallion = f.medallion AND t.hack_license = f.hack_license 
  AND t.pickup_datetime = f.pickup_datetime 
  AND t.passenger_count > 0 and t.passenger_count < 8 
  AND f.tip_amount >= 0 AND f.tip_amount <= 15 
  AND f.fare_amount >= 1 AND f.fare_amount <= 150 
  AND f.tip_amount < f.fare_amount AND t.trip_distance > 0 
  AND t.trip_distance <= 40 AND t.trip_distance >= 1
  AND t.trip_time_in_secs >= 30 AND t.trip_time_in_secs <= 7200 
  AND t.rate_code <= 5 AND f.payment_type in ('CSH','CRD')")
createOrReplaceTempView(trip_fareDF, "trip_fare")
SparkR::cache(trip_fareDF); SparkR::count(trip_fareDF);

###########################################
# SHOW REGISTERED TABLES
###########################################
head(SparkR::sql("show tables"))
```


#Feature engineering using SQL 
You can create new features using sQL statements. For example, you can use case statements to generate categorical features from coneunuous (numerical) ones.
```{r Feature engineering, message=FALSE, warning=FALSE, echo=TRUE}
###########################################
# CREATE FEATURES IN SQL USING CASE STATEMENTS
###########################################
trip_fare_feat <- SparkR::sql("SELECT 
    payment_type, pickup_hour, fare_amount, tip_amount, 
    passenger_count, trip_distance, trip_time_in_secs, 
  CASE
    WHEN (pickup_hour <= 6 OR pickup_hour >= 20) THEN 'Night'
    WHEN (pickup_hour >= 7 AND pickup_hour <= 10) THEN 'AMRush' 
    WHEN (pickup_hour >= 11 AND pickup_hour <= 15) THEN 'Afternoon'
    WHEN (pickup_hour >= 16 AND pickup_hour <= 19) THEN 'PMRush'
    END as TrafficTimeBins,
  CASE
    WHEN (tip_amount > 0) THEN 1 
    WHEN (tip_amount <= 0) THEN 0 
    END as tipped
  FROM trip_fare")

SparkR::cache(trip_fare_feat); SparkR::count(trip_fare_feat);
createOrReplaceTempView(trip_fare_feat, "trip_fare_feat")
head(trip_fare_feat, 3)
```
<hr>

#Data visualization
##Plots to inspect data and relationships between variables
For visualization, a small portion data will have to be sampled and brought into local memory as a data.frame object. R's plotting functions (e.g. from those in ggplot package) can then be applied to the data.frame for visualization.
```{r Exploration and visualization, message=FALSE, warning=FALSE, echo=TRUE, fig.width=10, fig.height=4}
###########################################
# SAMPLE SMALL PORTION OF DATA
###########################################
trip_fare_featSampled <- SparkR::sample(trip_fare_feat, withReplacement=FALSE, 
                                fraction=0.0001, seed=123)

###########################################
# CONVERT SPARK DF TO LOCAL DATA.FRAME IN MEMORY OF R-SERVER EDGE NODE
###########################################
trip_fare_featSampledDF <- as.data.frame(trip_fare_featSampled);

###########################################
# Generate HISTOGRAM OF TIP AMOUNT
###########################################
hist <- ggplot(trip_fare_featSampledDF, aes(x=tip_amount)) + 
  geom_histogram(binwidth = 0.5, aes(fill = ..count..)) + 
  scale_fill_gradient("Count", low = "green", high = "red") + 
  labs(title="Histogram for Tip Amount");

###########################################
# Generate Scatter Plot OF TRIP DISTANCE vs. TIP AMOUNT
###########################################
scatter <- ggplot(trip_fare_featSampledDF, aes(tip_amount, trip_distance)) + 
  geom_point(col='darkgreen', alpha=0.3, pch=19, cex=2) + 
  labs(title="Tip amount vs. trip distance");

###########################################
# Plot Histogram and Scatter Plot OF TIP AMOUNT Side by Side
###########################################
grid.arrange(hist, scatter, ncol=2)
```

##Advanced summarization using SQL and plotting:
We explore trips in NYC during rush and non-rush hours. This section shows more examples of SQL and advanced plotting and visualization, using ggmap. We plot the number of trips by day of the month, as well as number of trips on the NY City map during rush and non-rush hours.
```{r Advanced SQL and Visualization 1, message=FALSE, warning=FALSE, echo=TRUE, fig.width=10, fig.height=4}
###########################################
# GROUP TRIPS BY YEAR, MONTH, DAY, PAYMENT TYPE
###########################################
trip_stats_by_day <- SparkR::sql("select 
      year(pickup_datetime) as year, month(pickup_datetime) as month, 
      day(pickup_datetime) as day, payment_type, count(1) as trips 
      from fare 
      where payment_type in ('CSH','CRD')
      group by year(pickup_datetime), month(pickup_datetime), 
      day(pickup_datetime), payment_type")
tsbd <- as.data.frame(trip_stats_by_day)

###########################################
# PLOT NUMBER OF TRIPS BY DAY IN DEC 2013
###########################################
ggplot(data=tsbd, aes(day, trips)) + 
  geom_point(aes(color=payment_type)) + 
  geom_smooth(aes(color=payment_type))
```

#Modeling with SparkR
##Down-sample data for modeling
If a data-set is large, it may need to be down-sampled for modeling in reasonable amount of time. Here we used the <b>sample</b> function from SparkR to down-sample the joined trip-fare data. We then save the data in HDFS for use as input into the sparklyr modeling functions.
```{r Downsample data for training, message=FALSE, warning=FALSE, echo=TRUE}
###########################################
# SAMPLE DATA FOR MODELING
###########################################
trip_fare_featSampled <- sample(trip_fare_feat, withReplacement=FALSE, 
                                fraction=0.1, seed=123)
SparkR::cache(trip_fare_featSampled); SparkR::count(trip_fare_featSampled);
createOrReplaceTempView(trip_fare_featSampled, "trip_fare_featSampled")
```


##Partition data into train/test
```{r Partition data, message=FALSE, warning=FALSE, echo=TRUE, fig.width=4, fig.height=4}
###########################################
# PARTITION DATA INTO TRAIN-TEST USIN SQL
###########################################
dfrand <- SparkR::sql("SELECT *, RAND() as randnum from trip_fare_featSampled" );
trainDF <- SparkR::filter(dfrand, dfrand$randnum <= 0.7)
testDF <- SparkR::filter(dfrand, dfrand$randnum > 0.7)
```

##Create a SparkR::glm model
```{r Create model, message=FALSE, warning=FALSE, echo=TRUE, fig.width=4, fig.height=4}
########################################### 
## CREATE GLM MODEL
###########################################
model <- SparkR::glm(tip_amount ~ payment_type + pickup_hour + 
                    fare_amount + passenger_count + trip_distance + 
                    trip_time_in_secs + TrafficTimeBins, 
                    data = trainDF, family = "gaussian", 
                    epsilon = 1e-06, maxit = 5)
print (summary(model));
```

##Evaluate on a test set
```{r Evaluate on test set, message=FALSE, warning=FALSE, echo=TRUE, fig.width=4, fig.height=4}
########################################### 
## PREDICT ON TEST SET, AND EVALUATE ACCURACY
###########################################
predictions <- SparkR::predict(model, newData = testDF)
predfilt <- SparkR::select(predictions, c("label","prediction"))
predfilt_local <- SparkR::collect(predfilt)

# Evaluate and plot predictions (R-sqr)
Rsqr = cor(predfilt_local$label, predfilt_local$prediction)^2; Rsqr;

# Sample predictions for plotting
predictedSampled <- predfilt_local[base::sample(1:nrow(predfilt_local), 1000),]

# Plot predicted vs. actual values
lm_model <- lm(prediction ~ label, data = predictedSampled)
ggplot(predictedSampled, aes(label, prediction)) + 
  geom_point(col='darkgreen', alpha=0.3, pch=19, cex=2)   + 
  geom_abline(aes(slope = summary(lm_model)$coefficients[2,1], 
                  intercept = summary(lm_model)$coefficients[1,1]), 
              color = "red")
```


#Save data & model
##Save model for deployment
Partition data into train/test, and train a glm model and evaluate it.
```{r Persist model, message=FALSE, warning=FALSE, echo=TRUE}
###########################################
## SAVE MODEL
###########################################
modelPath =  file.path("/user/RevoShare/remoteuser/Models", "SparkGlmModel");
write.ml(model, modelPath) 
```

##Save joined and sampled data
Repartition the data in specific number of chunks and save
```{r Downsample and save data, message=FALSE, warning=FALSE, echo=TRUE}
###########################################
# REPARTITION DATA FOR SAVING
###########################################
trip_fare_featSampledRepartitioned <- repartition(trip_fare_featSampled, 10) # write.df below will produce this many files

###########################################
# SAVE DATAFRANE AS PARQUET file
###########################################
write.df(df=trip_fare_featSampledRepartitioned, 
         path=file.path(fullDataDir, "NYCjoinedParquetSubset"), 
         source="parquet", mode="overwrite")

```

#Uncache data and exit spark
```{r UNcache and exit , message=FALSE, warning=FALSE, echo=TRUE}
###########################################
# UNPERSIST CACHED DATA FRAME
###########################################
SparkR::unpersist(trip_fareDF)
SparkR::unpersist(trip_fare_feat)
SparkR::unpersist(trip_fare_featSampled)

###########################################
# STOP SPARKR CONTEXT
###########################################
sparkR.stop()
```

<br>
<hr>
<hr>
<br>

#Summary
The examples shown here can be adopted to fit other data exploration and modeling scenarios having different data-types or prediction tasks (e.g. classification).
